`mathml`: Translate R Expressions to MathML and LaTeX

### 2023-01-31 rj/mathml.Rmd
improve manuscript

The R\ extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
Since recently [@R, version 4.2], R\'s manual pages include support for
mathematical expressions [@Sarkar2022;@Viechtbauer2022], which already is a big
improvement. However, rules for the mapping of built-in language elements to
their mathematical representation are still lacking. So far, R\ expressions such
as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
such as \(P_{\mathrm{Bi}}(X \le k; N, p)\) require explicit Latex commands, that
is, `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for minimalistic use
cases, these commands are tedious to type in and their source code is hard to
read.

@@

The main feature of the package is that the same R\ expressions and equations
can be used for both mathematical typesetting and calculations. This saves time
and reduces mistakes, as will be illustrated below.

The paper is organized as follows. Section\ 2 presents the technical background
of the package, including the two main classes of Prolog rules for translating
R\ objects to mathematical expressions. Section\ 3 illustrates the main features
of the \CRANpkg{mathml} package, potential issues and workarounds using examples
from the day-to-day perspective of a user. Section\ 4 present a case study for
the use of the package in a scientific report. Section\ 5 concludes with a
discussion and ideas for further development.

@@

The translation of R\ expressions to mathematical output is achieved through a
Prolog interpreter from R\ package \CRANpkg{rolog} [@rolog]. Prolog is a classical logic
programming language with many applications in expert systems, computer
linguistics and symbolic artificial intelligence. The main strength of Prolog is
its concise representation of facts and rules for knowledge and grammar, as well
as its efficient built-in search engine for closed world domains. As it is
well-known, R\ is a statistical programming language for data analysis and
statistical modeling. Whereas Prolog is weak in statistical computation, but
strong in symbolic manipulation, the converse may be said for the R\ language.
The \pkg{rolog} package bridges this gap by providing an interface to a SWI-Prolog
distribution [@swipl] in R. The communication between the two systems is mainly
in the form of queries from R\ to Prolog, but two predicates allow Prolog to
ring back and evaluate terms in R.

For a first illustration of the \pkg{mathml} package, we consider the binomial
probability.

```{r}
term <- quote(pbinom(k, N, p))
term
```

The term is quoted to avoid its immediate evaluation (which would raise an error
anyway since the variables `k`, `N`, `p` have not yet been defined). Experienced
readers will remember that the quoted expression above is a short form for

```r
term <- call("pbinom", as.name("k"), as.name("N"), as.name("p"))
```

As is seen from the output above, the variable `term` is not assigned the result
of the calculation, but an R\ call
[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. Such a call
can eventually be evaluated with `eval()`,

```{r}
k <- 10
N <- 22
p <- 0.4
eval(term)
```

The R\ package \pkg{mathml} can now be used to render the call in MathML, that
is the dialect for mathematical elements on HTML webpages or in MathJax/LaTeX,
as shown below (some of the curly braces are not really needed in this simple
example, but are necessary in edge cases).

```{r}
library(mathml)
mathjax(term)
```

We can include the output in a RMarkdown document by specifying `results='asis'`
in the R\ code chunk, as is done in the next example. The
R\ function `mathout()` has been defined in this vignette; it invokes `mathml()`
for HTML output and `mathjax()` for LaTeX output.

```{r, results='asis'}
mathout(term)
```

@@

```prolog
math(pbinom(K, N, Pi), M)
 => M = fn(subscript('P', "Bi"), (['X' =< K] ; [N, Pi])).
```

Thus, the predicate `math/2` could be considered a "macro" that translates a
mathematical element (here, `pbinom(K, N, Pi)`) to a different mathematical
element, namely `fn(Name, (Args ; Pars))`. The low-level predicate `ml/3` is
used to convert these basic elements to MathML.

```prolog
ml(Flags, fn(Name, (Args ; Pars)), M)
 => ml(Flags, Name, N),
    ml(Flags, paren(list(op(;), [list(op(','), Args), list(op(','), Pars)])), X),
    M = mrow([N, mo(&(af)), X]).
```

The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
with `N` representing the name of the function and `X` its arguments and
parameters, enclosed in parentheses. A corresponding rule `jax/3` does the same
for MathJax/LaTeX. A list of flags can be used for context-sensitive
translation (see, e.g., the section on errors below).

@@

\pkg{mathml} is an R\ package for pretty mathematical representation of
R\ functions and objects in data analysis, scientific reports and interactive
web content. The currently supported features are listed below, roughly
following the order proposed by [@murrell2000].

@@

\pkg{mathml} handles the basic elements of everyday mathematical expressions,
such as numbers, Latin and Greek letters, multi-letter identifiers, accents,
subscripts, and superscripts.

@@

An expression such as `1 + -2` may be considered unsatisfactory from an
aesthetical perspective. It is correct R\ syntax, though, and is reproduced
accordingly, without the parentheses. Parentheses around negated numbers or
symbols can be added as shown above for `+ (-Inf)`.

@@

The package offers some support for different fonts as well as accents and
boxes etc. Internally, these decorations are implemented as identity functions,
so they can be introduced into R expressions without side-effects.

@@

For standard operators with known precedence, \pkg{mathml} is generally able to
detect if parentheses are needed; for example, parentheses are automatically
placed around `g + h` in the `nodot`-example. However, we note unnecessary
parentheses around `l + m` above. Thes parentheses are a consequence
of `quote(a^(b + c))` actually producing a nested R\ call of the
form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:

@@

For the present purpose, this feature is unfortunate because extra parentheses
around `b + c` are not needed. The preferred result is obtained by using the
functional form `quote('^'(k, l + m))` of the power, or curly braces as a
workaround (see `p + q` above).

@@

The example is a reminder that it is not possible to define the precedence of
custom operators in R, and that expressions with such operators are evaluated
strictly from left to right. Again, the problem can be worked around by the
functional form of the operator, or a curly brace to hide the parenthesis but
enforce the correct operator precedence.

@@

There is support for most functions from package _base_, with adequate use and
omission of parentheses.

@@

A few more examples are shown in Table \@ref(tab:base-stats), including
functions from `base` and `stats`.

@@

For self-written functions, matters are a bit more complicated. For a function
such as `g <- function(...) ...`, the name _g_ is not transparent to R, because
only the function body is represented, as illustrated by the redefinition
of `sign` below.

@@

As shown in the example, we can still display functions in the
form `head(x) = body` if we embed the object to be shown into a
call `"<-"(head, body)`.

The function body is generally a nested R\ call of the form `'{'(L)`, with `L`
being a list of commands (the semicolon, not necessary in R, is translated to a
newline). As illustrated in the example, \pkg{mathml} provides limited support
for control structures such as `if`.

@@

For mathematical typesetting in the form
of \(\int f(x)\, dx\), \pkg{mathml} needs to find out the name of the
integration variable. For that purpose, the underlying Prolog bridge provides a
predicate `r_eval/3` that calls R\ from Prolog. In the example above, this
predicate evaluates `formalArgs(args(sin))`, which returns the names of the
arguments of `sin`, namely, `x`.

@@

One of R's great features is the possibility to refer to function arguments by
their names, not only by their position in the list of arguments. At the other
end, Prolog does not have such a feature. Therefore, the Prolog handlers for
R\ calls are rather rigid, for example, `integrate/3` accepts exactly three
arguments in a particular order and without names, that
is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print the desired result.

@@

To improve the situation, \pkg{mathml} provides a simple hook that can be used
to replace elements (e.g., verbose variable names) of the code by concise
mathematical symbols, as illustrated in the example. To simplify notation,
the `quote` flag of `hook()` defaults to TRUE, and `hook()` uses non-standard
evaluation to unpack its arguments. If quote is FALSE, as shown above, the user
has to provide the quoted expressions. Care should be taken to
avoid recursive hooks such as `hook(s, s["A"])` that endlessly replace
the \(s\) from \(s_{\mathrm{A}}\) as
in \(s_{\mathrm{A}_{\mathrm{A}_{\mathrm{A}\cdots}}}\).

@@

For this purpose, the functions `mathml()` or `mathjax()` have an optional
argument `flags` which is a list with named elements. In this example, we use
this argument to tell \pkg{mathml} how to render such erroneous
expressions using the flag `error` which can be asis, highlight, fix, or ignore.
For more examples, see Table \@ref(tab:mistakes).

@@

## Customizing the package

```{r, results='asis'}
hook(successes, k)
hook(quote(Ntotal), quote(N), quote=FALSE)
hook(prob, pi)

term <- quote(dbinom(successes, Ntotal, prob))
mathout(term)
```

To improve the situation, \pkg{mathml} provides a simple hook that can be used
to replace elements (e.g., verbose variable names) of the code by concise
mathematical symbols, as illustrated in the example. To simplify notation,
the `quote` flag of `hook()` defaults to TRUE, and `hook()` uses non-standard
evaluation to unpack its arguments. If quote is FALSE, as shown above, the user
has to provide the quoted expressions. Care should be taken to
avoid recursive hooks such as `hook(s, s["A"])` that endlessly replace
the \(s\) from \(s_{\mathrm{A}}\) as
in \(s_{\mathrm{A}_{\mathrm{A}_{\mathrm{A}\cdots}}}\).

# A case study {#case-study}

@@

(One would typically use chunk option `echo=FALSE` to suppress the R code.)
In redundant stimuli, the activations from the channel-specific diffusion
processes add up, \(X_{\mathrm{AB}}(t) = X_{\mathrm A}(t) + X_{\mathrm B}(t)\),
hence the name, superposition. \(X_{\mathrm{AB}}(t)\) is again a Wiener process
with drift \(\mu_{\mathrm A} + \mu_{\mathrm B}\) and
variance \(\sigma^2_{\mathrm A} + \sigma^2_{\mathrm B}\). For the expected
first-passage time, we have

@@

# Conclusions

@@

The package \pkg{mathml} is available for R\ version (…) and later, and can be
easily installed using the usual `install.packages("mathml")`. The source code
of the package is found at https://github.com/mgondan/mathml.

>>>>

The R\ extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
Since recently [@R, version 4.2], R\'s manual pages include support for
mathematical expressions [@Sarkar2022;@Viechtbauer2022], which already is a big
improvement. However, rules for the mapping of built-in language elements to
their mathematical representation are still lacking. So far, R\ expressions such
as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
such as \(P_{\mathrm{Bi}}(X \le k; N, p)\) require explicit Latex commands
like `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for very basic use
cases, these commands are tedious to type in and their source code is hard to
read.

@@

The main feature of the package is that the same R\ expressions and equations
can be used for both mathematical typesetting and calculations. This saves time
and potentially reduces mistakes, as will be illustrated below.

The paper is organized as follows. We start with a description of the
technical background of the package, including the two main classes of rules for
translating R\ objects to mathematical expressions. The next section illustrates
the main features of the \CRANpkg{mathml} package, potential issues and their
workarounds using examples from the day-to-day perspective of a user. A case
study follows with a scientific report written with the help of the package. The
last section concludes with a discussion and ideas for further development.

@@

Similar to other high-level programming languages, R is homoiconic, that is,
R\ commands (i.e., R\ "calls") are, themselves, symbolic data structures that
can be created, parsed and modified. Because the default response of the
R\ interpreter is to evaluate a call and return its result, this property is not
transparent to the general user. There exist, however, a number of built-in
R\ functions (e.g., `quote()`, `call()` etc.) that allow to create R\ calls
which can be stored in regular variables and then, for example, be evaluated at
a later stage or in a specific environment [@Wickham2019]. The present package
includes a set of rules that translate such calls to a mathematical
representation in MathML and LaTeX. For a first illustration of the \pkg{mathml}
package, we consider the binomial probability.

````{r}
term <- quote(pbinom(k, N, p))
````

The term is quoted to avoid its immediate evaluation (which would raise an error
anyway since the variables `k`, `N`, `p` have not yet been defined). Experienced
R\ users will recognize that the expression is a short form for

````{r}
term <- call("pbinom", as.name("k"), as.name("N"), as.name("p"))
term
````

As is seen from the output, the variable `term` is not assigned the result of
the calculation, but an R\ call
[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. Such a call
can eventually be evaluated with `eval()`,

````{r}
k <- 10
N <- 22
p <- 0.4
eval(term)
````

The R\ package \pkg{mathml} can now be used to render the call in MathML or in
MathJax/LaTeX. MathML is the dialect for mathematical elements on HTML webpages,
whereas LaTeX is typically used for typesetting printed documents, as shown
below.

````{r}
library(mathml)
mathjax(term)
````

Some of the curly braces are not really needed in this simple example, but are
necessary in edge cases. We can include the output in a RMarkdown document by
specifying `results='asis'` in the R\ code chunk of RMarkdown, as is done in all
examples below. The R\ function `mathout()` has been defined in the preamble of
this article; it wraps a call to `mathml()` for HTML output and `mathjax()` for
LaTeX output.

````{r, results='asis'}
mathout(term)
````

For convenience, the translation of the R expressions is achieved through a
Prolog interpreter provided by another package 
R\ package \CRANpkg{rolog} [@rolog]. Prolog is a classical logic programming
language with many applications in expert systems, computer linguistics and
symbolic artificial intelligence. The strength of Prolog lies in its concise
representation of facts and rules for knowledge and grammar, as well as its
efficient built-in search engine for closed world domains. Whereas Prolog is
weak in statistical computation, but strong in symbolic manipulation, the
converse may be said for the R\ language. \pkg{rolog} bridges this gap by
providing an interface to a SWI-Prolog distribution [@swipl] in R. The
communication between the two systems is mainly in the form of queries from
R\ to Prolog, but two predicates allow Prolog to ring back and evaluate terms
in R.

@@

````prolog
math(pbinom(K, N, Pi), M)
 => M = fn(subscript('P', "Bi"), (['X' =< K] ; [N, Pi])).
````

Thus, the predicate `math/2` could be considered a "macro" that translates one
mathematical element (here, `pbinom(K, N, Pi)`) to another mathematical
element, namely `fn(Name, (Args ; Pars))`. The low-level predicate `ml/3` is
used to convert these basic elements to MathML.

````prolog
ml(fn(Name, (Args ; Pars)), M, Flags)
 => ml(Name, N, Flags),
    ml(paren(list(op(;), [list(op(','), Args), list(op(','), Pars)])), X, Flags),
    M = mrow([N, mo(&(af)), X]).
````

The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
with `N` representing the name of the function and `X` its arguments and
parameters enclosed in parentheses. A corresponding rule `jax/3` does the same
for MathJax/LaTeX. A list of flags can be used for context-sensitive
translation (see, e.g., the section on errors below).

@@

The currently supported R\ objects are listed below, roughly following the order
proposed by [@murrell2000].

@@

\pkg{mathml} handles the basic elements of everyday mathematical expressions,
such as integers, floating-point numbers, Latin and Greek letters, multi-letter
identifiers, accents, subscripts, and superscripts.

@@

An expression such as `1 + -2` may be considered aesthetically unsatisfactory.
It is correct R\ syntax, though, and is reproduced accordingly, without the
parentheses. Parentheses around negative numbers or symbols can be added as
shown above for `+ (-Inf)`.

@@

The package offers some support for different fonts as well as accents and
boxes etc. Internally, these decorations are implemented as identity functions,
so that they can be introduced into R expressions without side-effects.

@@

For standard operators with known precedence, \pkg{mathml} is generally able to
detect if parentheses are needed; for example, parentheses are automatically
placed around `g + h` in the `nodot`-example. However, we note unnecessary
parentheses around `l + m` above. These parentheses are a consequence
of `quote(a^(b + c))` actually producing a nested R\ call of the
form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:

@@

For the present purpose, this R\ feature is unfortunate because the extra
parentheses around `b + c` are not needed. The preferred result is obtained by
the functional form `quote('^'(k, l + m))` of the power, or curly braces as a
workaround (see `p + q` above).

@@

The example is a reminder that it is not possible to define the precedence of
custom operators in R, and that expressions with such operators are evaluated
strictly from left to right. Again, the problem can be worked around by the
functional form of the operator or a curly brace to hide the parenthesis, and,
at the same time, enforce the correct operator precedence.

@@

There is support for most functions from package `base`, with adequate use and
omission of parentheses.

@@

A few more examples are shown in Table \@ref(tab:base-stats), including
functions from `stats`.

@@

For self-written functions, matters are a bit more complicated. For a function
such as `g <- function(...) ...`, the name _g_ is not transparent to R, because
only the function body is represented. We can still display functions in the
form `head(x) = body` if we embed the object to be shown into a
call `"<-"(head, body)`.

@@

The function body is generally a nested R\ call of the form `'{'(L)`, with `L`
being a list of commands (the semicolon, not necessary in R, is translated to a
newline). The example also illustrates that \pkg{mathml} provides limited
support for control structures such as `if` that is internally represented
as `if(condition, action)`.

@@

For mathematical typesetting in the form
of \(\int f(x)\, dx\), \pkg{mathml} needs to find out the name of the
integration variable. For that purpose, the underlying Prolog bridge provides a
predicate `r_eval/2` that calls R\ from Prolog. This predicate is used to
evaluate `formalArgs(args(sin))` and returns the names of the arguments
of `sin()`, namely, _x_.

@@

One of R's great features is the possibility to refer to function arguments by
their names, not only by their position in the list of arguments. At the other
end, the Prolog handlers for R\ calls are rather rigid, for
example, `integrate/3` accepts exactly three arguments in a particular order and
without names, that is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print
the desired result.

@@

To improve the situation, \pkg{mathml} provides a simple hook that can be used
to replace elements (e.g., verbose variable names) of the code by concise
mathematical symbols, as illustrated in the example. To simplify notation,
`hook()` uses non-standard evaluation of its arguments. If the `quote` flag
of `hook()` is set to `FALSE`, the user has to provide the quoted expressions.
Care should be taken to avoid recursive hooks such as `hook(s, s["A"])` that
endlessly replace the \(s\) from \(s_{\mathrm{A}}\) as
in \(s_{\mathrm{A}_{\mathrm{A}_{\mathrm{A}\cdots}}}\).

@@

For this purpose, the functions `mathml()` or `mathjax()` have an optional
argument `flags` which is a list with named elements. In this example, we use
this argument to tell \pkg{mathml} how to render such erroneous
expressions using the flag `error` which can be "asis", "highlight", "fix", 
or "ignore". For more examples, see Table \@ref(tab:mistakes).

@@

# A case study

@@

One would typically use chunk option `echo=FALSE` to suppress the R code.

In redundant stimuli, the activation from the channel-specific diffusion
processes adds up, \(X_{\mathrm{AB}}(t) = X_{\mathrm A}(t) + X_{\mathrm B}(t)\),
hence the name, superposition. \(X_{\mathrm{AB}}(t)\) is again a Wiener process
with drift \(\mu_{\mathrm A} + \mu_{\mathrm B}\) and
variance \(\sigma^2_{\mathrm A} + \sigma^2_{\mathrm B}\). For the expected
first-passage time, we have

@@

# Conclusion

@@

The package \pkg{mathml} is available for R\ version 4.2 and later, and can be
easily installed using the usual `install.packages("mathml")`. The source code
of the package is found at https://github.com/mgondan/mathml.
